Archaic introgression and the distribution of shared functional variation under stabilizing selection

### 2023-12-14 doc/manuscript.tex
update doc, directional deserts

\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{color}

\usepackage{setspace}

@@

\DeclareMathOperator{\Var}{Var}

\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\E}{\mathbb{E}}

@@

@@

Genomic surveys of natural systems suggest that historical admixture among
diverged populations or closely related taxa commonly occurs
\citep{brandvain2014speciation,skoglund2015ancient,suvorov2022widespread} and
is potentially widespread in primate
\citep{tung2017contribution,sorensen2023genome} and hominin
\citep{wolf2018outstanding,peter2020100} evolution. Admixture may therefore be
an important driver of phenotypic and molecular variation and shape the genetic
basis of complex traits. In humans, archaic introgression involving
Neanderthals and Denisovans has attracted considerable attention, with efforts
to describe the historical processes leading to observed distributions of
introgressed DNA in present-day populations
\citep{prufer2014complete,villanea2019multiple,chen2020identifying} and its
contribution to quantitative traits
\citep{sankararaman2016combined,wei2023lingering}.

Once introduced through admixture, introgressed alleles may be selected for or
against. Some introgressed haplotypes were positively selected in modern humans
\citep{huerta2014altitude,racimo2017signatures,enard2018evidence,gower2021detecting},
possibly due to locally adaptive variation that provided fitness advantages as
humans encountered novel environments. Despite some cases of adaptive
introgression, introduced alleles were more likely to have been selected
against in modern humans \citep{harris2016genetic,juric2016strength}. Since
Neanderthal and Denisovan population sizes were relatively small for hundreds
of thousands of years, theory predicts they would have accumulated deleterious
variants at an increased rate. Introgressed haplotypes loaded with more
deleterious mutations would then have quickly been selected against after
admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
humans shows a reduction of Neanderthal-related ancestry in enhancers and
regulatory regions
\citep{petr2019limits,telis2020selection,yermakovich2023long}. These
``deserts'' of Neanderthal ancestry support the hypothesis that introgressed
functional alleles were selected against
\citep{sankararaman2014genomic,sankararaman2016combined}.

There is growing genetic evidence that early \emph{H. sapiens} reciprocally
contributed to Neanderthal genomes
\citep{kuhlwilm2016ancient,hubisz2020mapping,harris2023diverse}. This
human-to-Neanderthal introgression occured tens or hundreds of thousands of
years prior to Neanderthal introgression in humans during modern human global
dispersal around 60 ka. This is supported by ``near modern'' \emph{H. sapiens}
outside of African around 120--100ka or earlier
\citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
with Neanderthals and providing opportunities for early contacts. While
estimates of the genomic contribution of early \emph{H. sapiens} to
Neanderthals vary, around $6\%$ of later Neanderthal genomes may have been
contributed through introgression from humans to Neanderthals
\citep{harris2023diverse}. If \emph{H. sapiens}-related haplotypes carried
fewer deleterious alleles due to their larger long-term effective population
size, human-introgressed DNA would have been favored in Neanderthal genomes.
The replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports such a model of post-admixture selection in the Neanderthal
lineage \citep{posth2017deeply,petr2020evolutionary}.

Models for selection against introgressed alleles are often based on load
arguments (in particular, differences in the rate of accumulation of
unconditionally deleterious variation in populations of different sizes) or
hybrid incompatibilities \citep{muller1942isolating}. Such arguments, founded
in population genetics theory, rarely take into account selection operating on
phenotypic variation or the relationship between genetic and phenotypic
variation. Many phenotypic traits are thought to be under stabilizing selection
\citep{sanjak2018evidence,sella2019thinking}, including gene regulation
\citep{gilad2006natural,hodgins2015gene,price2022detecting}. Because some of
the strongest signals of selection against Neanderthal-introduced alleles are
in regulatory regions \citep{sankararaman2014genomic}, stabilizing selection on
quantitative traits is particularly relevant to the dynamics of functional
genetic variation after introgression.

Stabilizing selection acts to maintain the phenotypic distribution of a trait
near some optimum, which is achieved by reducing phenotypic variation
(Figure~\ref{fig:stab-sel}A). When the mean phenotype of the population is close
to the phenotypic optimum, classical models predict that alleles contributing
to such a trait are subject to underdominant selection, i.e., selection against
the minor allele at a locus \citep{robertson1956effect}. This has proven to be
a useful model for understanding genetic architectures of traits under
stabilizing selection in single-population settings
\citep[e.g.,][]{keightley1988quantitative, simons2018population,
hayward2022polygenic}.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Stabilizing selection and additive genetic variance in one population.}
        \aprcomment{See comparisons to simulations assuming free recombination in
        Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
        \aprcomment{$w$ and $f$ need to match text}
        \aprcomment{Add the SHC model, and the ``vary SD'' fig as a panel here}
        \aprcomment{First figure should just be two left panels. Combine two
        right panels to figure 2.}
    }
    \label{fig:stab-sel}
\end{figure}

In diverged populations, genetic variation contributing to a trait under
stabilizing selection has a higher rate of turnover compared to neutrally
evolving loci \citep{yair2022population}. We might therefore expect an
accumulation of fixed differences in the \emph{H. sapiens} and Neanderthal
lineages at loci contributing to such a trait, even when the mean phenotype in
each population remains close to the same trait optimum. When a derived allele
at high frequency in one population is introduced to another population in
which is was previously absent, it will be at low frequency and subsequently
selected against. Likewise, if the ancestral allele is reintroduced to a
population fixed for a derived allele, the ancestral allele will be at low
frequency and will be selected against. In either case we should expect
selection against the introgressed allele, whether ancestral or derived and
regardless the historical relative sizes of the populations involved. This
prediction contrasts with the population-genetics ``load'' model, in which
haplotypes with fewer deleterious variants (such as those from the population
with larger historical size) are favored after introgression in either
direction.

\subsection*{Model}

We consider a polygenic trait for which an individual's (additive) genetic
value is the sum over all effects of alleles in their genome, so that for
individual $i$, \(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size
of the derived allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their
genotype at that locus (i.e., the number of alleles they carry at that locus).
With no linkage between trait-affecting loci, the expected additive genetic
variance is \(V_A = \sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele
frequency at locus $l$. We ignore dominance and epistatic effects, so that
\(V_G = V_A\).  We further ingore environmental effects
\citep{simons2018population}, so that the phenotypic variance \(V_P=V_A\).

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $O$ (typally set to zero), and we assume a Gaussian fitness function
(Figure~\ref{fig:stab-sel}A) so that relative fitness is given by \(f(G_i | O,
V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $V_S$ can be interpreted as the strength
of selection on the trait, where larger $V_S$ implies weaker selection. For a
population with mean phenotype at (or very close to) the optimum, the mean
fitness of the population (assuming a roughly normal distribution of phenotypic
values in the population) is \[\bar{w} \approx \int_{-\infty}^\infty f(G | 0,
V_S) \mathcal{N}(0, V_G) dG = \left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\] Thus,
as the genetic variance increases, mean fitness among individuals in the
population decreases.

\subsubsection*{Mutation rates, effect sizes and genetic variance}

@@

Mutations are not expected to each have the same effect size $a$, but rather
follow some distribution. Throughout, we assume effect sizes are drawn from a
normal distribution with mean 0 and given variance $V_M$. When population sizes
or effect sizes are small, drift dominates the dynamics of $V_G$, and at steady
state \citet{lande1976natural} found \[V_G \approx 4 N_e \mu V_M.\]
Interpolating between the selection- and drift-dominated regimes,
\begin{align}\label{eq:SHC}
    V_G & \approx \frac{4 \mu V_S}{1 + \frac{VS}{N_e V_M}}.
\end{align}
This, the ``stochastic
house-of-Cards'' (SHC) approximation (Figure~\ref{fig:stab-sel}B), was given by
\citet{burger1989much} and is discussed in detail in \citet[][Ch.
28]{walsh2018evolution}.

\subsubsection*{Approximating allelic dynamics via underdominance}

We model the distribution of allele frequencies underlying the trait using
the approximation that alleles are subject to symmetric underdominant
selection \citep{robertson1956effect,keightley1988quantitative}.
For an allele with (heterozygote) additive effect size $a$,
the selection coefficient
\[s\approx \frac{a^2}{2(V_S + V_G)} \approx \frac{a^2}{2V_S},\]
if $V_G \ll V_S$ \citep[e.g.,][]{simons2018population}. Allelic dynamics can
be modeled using the diffusion approximation, where the expected change
in mean allele frequency per generation is
\[\E[M_{\delta_p}] \approx -s p(1-p)(1-2p),\]
and the expected change in variance of the allele frequency per generation is
\[\E[V_{\delta_p}] \approx \frac{1}{4N_e}p(1-p).\]
%From this, because $s$ is always positive for any $a\not=0$, we see that
%selection pushes allele frequencies to zero if $p<1/2$ and to one if $p>1/2$.
%$p=1/2$ is an unstable equilibrium.

We extend the moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017inferring} to include underdominance with given
selection coefficient $s$ (\(=a^2/2(V_S+V_G)\)). The contribution of alleles
with effect size $a$ to the total genetic variance $V_G$ is then found by
computing the expected pairwise diversity from the SFS (with sample size $n$,
$\Phi_n$), as
\[V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da.\]
Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. Then
assuming a normal distribution of effect sizes for new mutations, the total
genetic variance is
\[V_G=\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]

If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
$s=a^2/2V_S$ leads to lower estimates of additive genetic variance
(Figure~\ref{fig:toy-admixture}B). When mutation rates are large so that $V_G$
is not small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of
$V_G$ that closely match simulations assuming free recombination between loci
(Figure~\ref{fig:toy-admixture}X,Y). Because $V_G$ can change over time, this
means that $s$ is no longer constant, but can change due to factors such as
non-constant demography that increase or reduce $V_G$.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/reciprocal_admixture.pdf}
    \caption{
        \textbf{Predicting the genetic variance of a trait under stabilizing
        seletion and non-steady state demography.}
        \aprcomment{Make it in generations?}
        \aprcomment{Right two panels should go to supplement, move single-pop
        with size change from figure 1 to here.}
    }
    \label{fig:toy-admixture}
\end{figure}

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration and admixture. Here, we
consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In a simple toy model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, one remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{fig:toy-admixture}X,Y).

The second model is meant to more closely resemble inferred human-Neanderthal
history, in which the ancestral population of size $N_e=10\,000$ splits into
the human and Neanderthal branches. At 250ka, an early human-to-Neanderthal
introgression contributes $5\%$ ancestry to Neanderthals. The human branch
shrinks to size $1\,000$ 60ka, followed by exponential growth to $20\,000$ at
present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked and
expanding human population at 50ka, after which they go extinct
(Figure~\ref{fig:neand-to-human}A).

@@

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/neanderthal_admixture.pdf}
    \caption{
        \textbf{}
        \aprcomment{Show total/introgressed/non-introgressed in both directions
        with both $\sigma_M$.}
    }
    \label{fig:neand-to-human}
\end{figure}

@@

In the preceding sections, we find that approximating the dynamics of
trait-affecting alleles using an underdominance model
\citep{robertson1956effect} provides an excellent approximation of $V_G$ in
complex demographic scenarios. However, this relies on an absence of linkage
between trait-affecting alleles. The inclusion of linkage can lead to
noticeable effects on expected $V_G$, so that Equation~\ref{eq:SHC} slightly
overestimates $V_G$ at steady state
\citep{burger1989much,burger1994distribution,walsh2018evolution}.

>>>>

\usepackage[letterpaper,margin=1in]{geometry}
\usepackage{lineno}   %% <- no mathlines option
\usepackage{amsmath}  %% <- after lineno
\usepackage{etoolbox} %% <- for \cspreto, \csappto
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{color}
\usepackage{url}

\usepackage{setspace, caption}
\captionsetup{font=onehalfspacing} % Spacing in float captions

@@

\DeclareMathOperator{\Var}{Var}
\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
\newcommand{\E}{\mathbb{E}}
\usepackage{xspace}
\newcommand{\moments}{\texttt{moments}\xspace}
\newcommand{\fwdpy}{\texttt{fwdpy11}\xspace}
\newcommand{\tskit}{\texttt{tskit}\xspace}

%% Patch 'normal' math environments:
\newcommand*\linenomathpatch[1]{%
  \cspreto{#1}{\linenomath}%
  \cspreto{#1*}{\linenomath}%
  \csappto{end#1}{\endlinenomath}%
  \csappto{end#1*}{\endlinenomath}%
}

\linenomathpatch{equation}
\linenomathpatch{gather}
\linenomathpatch{multline}
\linenomathpatch{align}
\linenomathpatch{alignat}
\linenomathpatch{flalign}

%\linenumbers

@@

    Many phenotypic traits are under stabilizing selection, which maintains
    individual phenotypic values near some optimum. Classical theory shows that
    at loci contributing to the genetic variance of a trait under stabilizing
    selection, the minor allele is selected against via symmetric
    underdominance. Most studies of the dynamics of the genetic architecture
    and variance of a trait under stabilizing selection have focused on single
    populations. However, natural populations are typically structured in some
    way, and admixture and introgression are common occurances, in particular
    in human evolution. Because stabilizing selection results in selection
    against the minor allele at a trait-affecting locus, the alleles from the
    minor parental ancestry will be selected against after admixture. Here, we
    show that the site frequency spectrum can be used to model the genetic
    architecture of such traits, allowing for complex multi-population
    histories including admixture. We develop a simple model with linkage to a
    neutral locus to show that introgressed ancestry is expected to be depleted
    around trait-affecting alleles. Using this model and individual-based
    simulations, we show that introgressed ancestry deserts tend to form around
    such loci. When introgression between two diverged populations occurs in
    both directions, as has been inferred between humans and Neanderthals,
    introgressed ancestry deserts will often be shared across populations.
    Stabilizing selection for a shared phenotypic optimum may therefore explain
    recent observations in which regions of depleted human-introgressed
    ancestry in the Neanderthal genome overlap with Neanderthal-ancestry
    deserts in humans.

@@

Genomic surveys of natural systems show that historical admixture among
diverged populations or closely related taxa commonly occurs
\citep{brandvain2014speciation, skoglund2015ancient, suvorov2022widespread} and
is potentially widespread in primate \citep{tung2017contribution,
sorensen2023genome} and hominin \citep{wolf2018outstanding, peter2020100}
evolution. Admixture may therefore be an important driver of phenotypic and
molecular variation and shapes the genetic basis of complex traits. In humans,
archaic introgression involving Neanderthals and Denisovans has attracted
considerable attention, including efforts to describe the historical processes
leading to observed distributions of introgressed DNA in present-day
populations \citep{prufer2014complete, villanea2019multiple,
chen2020identifying} and the contribution of introgressed variation to
quantitative traits \citep{sankararaman2016combined, wei2023lingering}.

Once introduced through admixture, introgressed alleles may be selected for or
against. Some introgressed haplotypes were likely positively selected in modern
humans \citep{huerta2014altitude, racimo2017signatures, enard2018evidence,
gower2021detecting}, possibly due to locally adaptive variation that provided
fitness advantages as humans encountered novel environments. Despite some cases
of adaptive introgression, mostly introduced alleles were likely to have been
selected against in humans \citep{harris2016genetic, juric2016strength}. Since
Neanderthal and Denisovan population sizes were relatively small for hundreds
of thousands of years, theory predicts they would have accumulated fixed
deleterious variation at an increased rate. Introgressed haplotypes loaded with
more deleterious mutations would then have quickly been selected against after
admixture. Mapping the distribution of Neanderthal-introgressed haplotypes in
humans shows a reduction of Neanderthal-related ancestry in coding and
regulatory regions \citep{petr2019limits, telis2020selection,
yermakovich2023long}. These ``deserts'' of Neanderthal ancestry support the
hypothesis that introgressed functional alleles were selected against
\citep{sankararaman2014genomic, sankararaman2016combined}.

There is growing genetic evidence that early \emph{H. sapiens} reciprocally
contributed to Neanderthal genomes \citep{kuhlwilm2016ancient,
hubisz2020mapping, harris2023diverse}. Such introgression would have occurred
tens or hundreds of thousands of years prior to Neanderthal introgression in
humans during the global dispersal of modern humans around 60 ka. This is
supported by ``near modern'' \emph{H. sapiens} outside of African around
120--100 ka or earlier \citep{schwarcz1988esr, grun2005u, beyer2021climatic},
potentially overlapping with Neanderthals and providing opportunities for early
contacts. While estimates of the genomic contribution of early \emph{H.
sapiens} to Neanderthals vary, around $6\%$ of later Neanderthal genomes may
have been contributed through introgression from humans to Neanderthals
\citep{harris2023diverse}. If \emph{H. sapiens}-related haplotypes carried
fewer deleterious alleles due to their larger long-term effective population
size, human-introgressed DNA would have been favored in Neanderthal genomes.
The replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports this model of post-admixture selection in the Neanderthal
lineage \citep{posth2017deeply, petr2020evolutionary}.

Models for selection against introgressed alleles are often based on load
arguments (in particular, differences in the rate of accumulation of
unconditionally deleterious variation in populations of different sizes) or
hybrid incompatibilities \citep{muller1942isolating}. These arguments, founded
in population genetics theory, rarely take into account selection operating on
phenotypic variation or the relationship between genetic and phenotypic
variation. Many phenotypic traits are thought to be under stabilizing selection
\citep{sanjak2018evidence, sella2019thinking}, including gene regulation
\citep{gilad2006natural, hodgins2015gene, price2022detecting}. Because some of
the strongest signals of selection against Neanderthal-introduced alleles are
in regulatory regions \citep{sankararaman2014genomic}, stabilizing selection on
quantitative traits is particularly relevant to the dynamics of functional
genetic variation after introgression.

Stabilizing selection acts to maintain the phenotypic distribution of a trait
near some optimum, which is achieved by reducing phenotypic variation
(Figure~\ref{fig:stab-sel}A). When the mean phenotype of the population is
close to the phenotypic optimum, classical models predict that trait-affecting
alleles are subject to underdominant selection, i.e., selection against the
minor allele at the locus \citep{robertson1956effect}. This has proven to be a
useful model for understanding genetic architectures of traits under
stabilizing selection in single-population settings
\citep[e.g.,][]{keightley1988quantitative, simons2018population,
hayward2022polygenic}.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/stab_sel.pdf}
    \caption{
        \textbf{Additive genetic variance under stabilizing selection.}
        (A) Stabilizing selection acts to maintain phenotypic values of individuals
        in the population near some optimum. Throughout, we assume a Gaussian fitness
        function.
        (B) With low mutational variance ($V_M$), the expected additive genetic
        variance ($V_G$) is proportional to the population-scaled mutation rate.
        When $V_M$ is large so that mutational effects can be strong, $V_G$ is
        independent of the mutation rate. The stochastic house-of-cards model
        (Eq.~\ref{eq:SHC}) interpolates these regimes, assuming steady-state
        dynamics \citep{burger1989much}. Expected $V_G$ computed using an SFS
        \citep[using \moments,][]{jouganous2017inferring} matches
        simulations assuming no linkage between loci affecting the trait, which
        align closely the stochastic house-of-cards approximation. Here, the
        mutation rate $\mu=0.01$ (per haploid), $V_S=1$, and $N_e=10,000$.
        Mutation effects are normally distributed with mean zero and variance
        $V_M$.
    }
    \label{fig:stab-sel}
\end{figure}

In diverged populations, the genetic variation contributing to a trait under
stabilizing selection has a higher rate of turnover compared to neutrally
evolving loci \citep{yair2022population}. We might therefore expect a rapid
divergence of trait architecture and an accumulation of fixed differences in
the \emph{H. sapiens} and Neanderthal lineages at trait-contributing loci, even
when the mean phenotype in each population remains close to the same trait
optimum. When a derived allele at high frequency in one population is
introduced to another population in which is was previously absent, it will be
at low frequency (if the admixture proportion is low) and subsequently selected
against. Likewise, if the ancestral allele is reintroduced to a population
fixed for a derived allele, the ancestral allele will be at low frequency and
will be selected against. In either case we should expect selection against the
introgressed allele, whether ancestral or derived and regardless the historical
relative sizes of the populations involved. This prediction contrasts with the
population-genetics ``load'' model, in which haplotypes with fewer deleterious
variants (such as those from the population with larger historical size) are
favored after introgression in either direction.

In this article, we show that admixture between diverged populations results in
selection against the minor parental ancestry at trait-affecting loci. We
develop a numerical approach based on the site frequency spectrum to predict
genetic variation under complex demographic scenarios, and which we use to
partition predicted trait heritability by introgressed vs. non-introgressed
variation. Using simulations with linkage, we demonstrate that deserts of
introgressed ancestry form around trait-contributing loci. When gene flow
occurs bidirectionally, such deserts will tend to overlap in location in each
population. As we argue in the Discussion, stabilizing selection on shared
trait optima provides an explanation for the overlap of introgressed ancestry
deserts in human and Neanderthal genomes after reciprocal introgression,
recently observed in \citet{harris2023diverse}.

\section*{Model and Methods}

We consider a polygenic trait for which an individual's additive genetic value
is the sum over all effects of alleles in their genome. For individual $i$,
\(G_i = \sum_l g_{l, i} a_l\), where $a_l$ is the effect size of the derived
allele at locus $l$, and \(g_{l,i}\in\{0,1,2\}\) is their genotype at that
locus (i.e., the number of alleles they carry at that locus). With no linkage
between trait-affecting loci, the expected additive genetic variance is \(V_A =
\sum_l 2p_l(1-p_l)a_l^2\), where $p_l$ is the allele frequency at locus $l$. We
ignore dominance and epistatic effects, so that \(V_G = V_A\). We further
ingore environmental effects \citep{simons2018population}, so that the
phenotypic variance \(V_P=V_G\).

Stabilizing selection acts to reduce phenotypic variation around the optimum
value $O$ (typally set to zero), and we assume a Gaussian fitness function
(Figure~\ref{fig:stab-sel}A) so that relative fitness is given by \(f(G_i | O,
V_S) = \exp{(-(G_i - O)^2 / 2 V_S)}\). $\frac{1}{V_S}$ is interpreted as the
strength of selection on the trait, so larger $V_S$ implies weaker selection.
For a population with mean phenotype at (or very close to) the optimum, the
mean fitness of the population (assuming a roughly normal distribution of
phenotypic values in the population) is
\[\bar{w} \approx \int_{-\infty}^\infty
f(G | 0, V_S) \mathcal{N}(0, V_G) dG =
\left(\frac{V_S}{V_S+V_G}\right)^{1/2}.\]
Thus, as the genetic variance increases, mean fitness among individuals in the
population decreases.

\subsection*{Mutation rates, effect sizes and genetic variance}

@@

Mutations are not expected to each have the same effect size $|a|$, but rather
follow some distribution. Here, when mutation effects are not constant, we
assume effect sizes are drawn from a normal distribution with mean 0 and given
variance $V_M$. When population sizes or effect sizes are small, drift
dominates the dynamics of $V_G$, and at steady state \citet{lande1976natural}
found \[V_G \approx 4 N_e \mu V_M.\] Interpolating between the selection- and
drift-dominated regimes,
\begin{align}\label{eq:SHC}
    V_G & \approx \frac{4 \mu V_S}{1 + \frac{V_S}{N_e V_M}}.
\end{align}
This, the ``stochastic
house-of-Cards'' (SHC) approximation (Figure~\ref{fig:stab-sel}B), was given by
\citet{burger1989much} and is discussed in detail in \citet[][Ch.
28]{walsh2018evolution}.

\subsection*{Approximating allelic dynamics via underdominance}

As initially shown by \citet{robertson1956effect} \citep[see
also,][]{keightley1988quantitative, simons2018population}, when the mean
phenotype of the population is close to the optimal phenotypic value,
stabilizing selection results in symmetric underdominant selection on alleles
contributing additively to the trait. Thus, for an allele at frequency $p$ with
selection coefficient $s$, the expected change in $p$ over one generation is
\begin{equation}
    \E[\Delta p] = -sp(1-p)(1-2p).
\end{equation}
The selection coefficent $s$ depends on the strength of selection on the trait
$V_S$ and the effect size $a$, as 
\begin{equation}
    s\approx \frac{a^2}{2(V_S+V_G)} \approx \frac{a^2}{2V_S}.
\end{equation}
Details are shown in Appendix~\ref{sec:underdominance}.

\subsubsection*{Computing expected genetic variance from the SFS}

We model allelic dynamics using the diffusion approximation, where the expected
change in mean allele frequency per generation is
\[\E[M_{\delta_p}] = E[\Delta p] \approx -s p(1-p)(1-2p),\]
as shown above, and the expected change in variance of the allele frequency per
generation is
\[\E[V_{\delta_p}] \approx \frac{1}{2N_e}p(1-p).\]
Because $s$ is always positive for any $a\not=0$, we see that selection pushes
allele frequencies to zero if $p<1/2$ and to one if $p>1/2$, resulting in
symmetric underdominance.

We extend the \moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017inferring} to include underdominance with given
selection coefficient $s$ (Appendix~\ref{sec:moments-underdominance}). The
contribution of alleles with effect size $a$ to the total genetic variance
$V_G$ is found by computing the expected pairwise diversity from the SFS (with
sample size $n$, denoted $\Phi_n$), as \[V_{G,a} \approx 2a^2\sum_{j=1}^{n-1}
\frac{j(n-j)}{n(n-1)} \Phi_n(j|a,\mu_a) da.\] Here, $\mu_a$ is the mutation
rate of alleles with effect size $a$. Then assuming a normal distribution of
effect sizes for new mutations, the total genetic variance is \[V_G \approx
\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]

If $V_G$ is non-negligible compared to $V_S$, ignoring $V_G$ and using
$s=a^2/2V_S$ leads to deflated estimates of additive genetic variance
(Figure~\ref{fig:supp-one-pop}). When mutation rates are large so that $V_G$ is
not small compared to $V_S$, using $s=a^2/2(V_S+V_G)$ provides estimates of
$V_G$ that closely match simulations assuming free recombination between loci
(Figures~\ref{fig:toy-admixture} and
\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}). Because $V_G$ can vary over
time, this means that $s$ is no longer constant and can change due to factors
such as non-constant demography that increase or reduce $V_G$.

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Population size changes and additive genetic variance.}
        We used \moments to track expected genetic variance of a trait
        under stabilizing selection as the population chages size. (A) The
        population goes through a 10-fold size reduction, followed by a recovery.
        (B) All mutations had effect sizes $\pm0.01$, $0.04$, or $0.1$, with equal
        probabilitiy of being trait increasing or decreasing. Depending on the
        selection coefficient ($s\approx a^2/2V_S$) compared to $N_e$, $V_G$
        could increase or decrease after a sudden population size change.
        Predictions using \moments match simulations with free recombination
        between trait-affeting loci
        (Figures~\ref{fig:one-pop-0.01}--\ref{fig:one-pop-0.1}).
    }
    \label{fig:one-pop}
\end{figure}

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration and admixture. Here, we
consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In a simple toy model (which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence) a population of size $N_e=10{,}000$ splits into two, one remaining
size $10{,}000$ and the other shrinking to size $1{,}000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{fig:toy-admixture}A,B).

The second model is meant to more closely resemble inferred human-Neanderthal
history, in which the ancestral population of size $N_e=10{,}000$ splits into the
human and Neanderthal branches. At 250ka, an early human-to-Neanderthal
introgression contributes $5\%$ ancestry to Neanderthals. The human branch
shrinks to size $1{,}000$ 60ka, followed by exponential growth to size $20,000$
at present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked
and expanding human population at 50ka, after which they go extinct
(Figure~\ref{fig:human-neand-h2}A).

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/reciprocal_admixture.pdf}
    \caption{
        \textbf{Genetic variance of a trait under stabilizing seletion with
        multi-population demography.}
        (A,B) Two simple demographic models, in which Deme 1 has size 10,000
        and Deme 2 has size 1,000. In each scenario, we allow $5\%$ admixture
        from one deme to the other after begin isolated for 20,000 ($2N_e$)
        generations.
        (C,D) We compared predicted (additive) $V_G$ from the site frequency
        spectra (using \moments) to simulations
        without linkage between trait-affecting loci. In Deme 2, $V_G$ decreases
        after divergence and population size reduction. After admixture in both
        directions, $V_G$ increases and then decays again toward levels at steady
        state. Predictions from \moments closely match observed $V_G$ in
        simulations.
        Here, mutations were drawn from a normal distribution of fitnesses
        with $V_M=0.0025$.
        Other parameters: $\mu=0.025$, $V_S=1$, and the optimal phenotype remained
        the same in each population.
        (E,F) Using \moments, we partitioned additive contributions to $V_G$ by
        mutations that were previously segregating in the focal population,
        introgressed variants (either derived or reintroduced ancestral alleles),
        and by new mutations since the time of admixture. In both demographic
        scenarios, introgressed variants contribute a substantial proportion to
        $V_G$, though it is primarily composed of introduced derived alleles when
        admixture is from the small to the large population, while primarily
        reintroduced ancestral alleles under the reverse direction of gene flow.
        In both cases, $V_G$ is increasingly due to new mutations and the genetic
        architecture of the trait turns over with time since admixture.
    }
    \label{fig:toy-admixture}
\end{figure}

\subsubsection*{Simulations with free recombination}

We compare our \moments-based predictions for $V_G$ in non-equilibrium settings
to simulations assuming linkage equilibrium as well as individual-based
simulations with recombination (next section). For both, we allow mutations
to arise at rate $mu$ per haploid genome copy per generation. These are drawn
from some distribution of mutation effects, which then contribute to individual
trait values.

To simulate allele frequency changes under free recombination (assuming linkage
equilibrium between all trait-affecting alleles at all times), for each focal
segregating allele we integrate over possible genetic backgrounds contributed
by all other segregating alleles. We make the assumption that many alleles
contribute to the trait, so the variance in genetic backgrounds is normally
distributed around the mean genetic value $E_{\tilde{G}}=\sum_l 2p_l a_l$ with
variance $V_{\tilde{G}}=\sum_l 2p_l(1-p_l)a_l^2$, where the sums omit the focal
locus. The expected change in frequency is then computed using the approach
outlined in Appendix~\ref{sec:underdominance} (without taking the first-order
Taylor series approximation for the exponentials). Allele counts in the next
generation are then binomially sampled with parameter $p+\E[\Delta p]$
independently for each allele.

\subsubsection*{Individual-based simulations with linkage}

To include the effects of linkage between multiple selected alleles or selected
and neutral alleles, we used \fwdpy \citep{thornton2019polygenic} to run
Wright-Fisher simulations under the same demographic models considered above.
In these simulations, we consider large (1 Morgan, or 100 Mb with a per-base
recombination rate of $10^-8$) chromosomes with a uniform recombination
landscape and specify per-haploid mutation rates. Mutations may be uniformly
distributed across the chromosome, as in Figures~\ref{fig:linkage} and
\ref{fig:supp-low-VM}--\ref{fig:supp-high-VM}, or they may fall within
functional regions, as in Figures~\ref{fig:deserts} and
\ref{fig:deserts-directional}. In these simulations, such regions are centered
2 Mb apart and are 100kb in size, so that there are 50 evenly spaced regions
across the chromosome. Mutation effect sizes are drawn either as constant
values $\pm a$, or from a normal distribution with given mutational variance
$V_M$.

In these simulations, we track the empirical phenotypic variance (since there
is no simulated environmental effect, this is equivalent to $V_G$) in each
population each generation. We measure the effects of linked selection on
(neutral) introgressed ancestry using \tskit to analyze genealogical
information \citep{ralph2020efficiently}. To measure the reduction in
introgressed ancestry around introgressed variants (as shown in
Figure~\ref{fig:linkage}), we find each locus with a fixed difference between
the parental populations at the time of admixture by preserving the generation
immediately preceding admixture. For such a locus, for each sample we determine
which (preserved) parental population its ancestry traces to at varying
distances from the selected locus. Ancestry proportions are then averaged over
each fixed difference. To measure proportions of introgressed ancestry or
probabilities of observed ancestry deserts (as in Figure~\ref{fig:deserts}), we
take the same approach to average ancestral source population proportions in 50
kb windows. Ancestry deserts are defined as any such window with no ancestry
inherited from the source population of introgression events.

\subsection*{Data and code availability}

All code to run analyses, create figures, and compile this manuscript are available
at \url{https://github.com/apragsdale/neanderthal_stabilizing_selection}.

@@

In the preceding sections, we find that approximating the dynamics of
trait-affecting alleles using an underdominance model
\citep{robertson1956effect} provides an excellent approximation of $V_G$ in
complex demographic scenarios. However, this relies on an absence of linkage
between trait-affecting alleles. The inclusion of linkage can lead to
noticeable effects on expected $V_G$, so that Equation~\ref{eq:SHC} differs
from observed $V_G$ at steady state
\citep{burger1989much, burger1994distribution, walsh2018evolution}.
